Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-09-26 paper/paper.tex
changes to enable  merge

\author{Fan~Cheng, Anastasios~Panagiotelis, Rob J~Hyndman}
\addresses{\textbf{Fan Cheng}\newline
Monash University
\newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
\textbf{Anastasios Panagiotelis}\newline
The University of Sydney
\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
\textbf{Rob J Hyndman}\newline
Monash University
\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
}

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of \(\mathbb{R}^d\). Applications of kernel density estimation {[}KDE; \textcite{Parzen1962-gt}; \textcite{Chen2017-dw}{]} include finding hot spots of traffic network in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac}, prime prediction via Twitter messages \autocite{Gerber2014-tq}, and so on. However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than \(\mathbb{R}^d\), kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted. \textcite{Pelletier2005-vu} propose a kernel density estimator based on the Riemannian geodesic distance of the manifold but it is only applicable when the underlying manifold is known. Manifold learning algorithms could be applied to reduce the dimension and get an approximation of the manifold, but different manifold learning algorithms could induce different distortions of the same manifold. Therefore, we propose a distortion-corrected kernel density estimator for Riemannian manifolds embedded in more than \(\mathbb{R}^d\). Our estimator could be applied to any reasonable manifold learning embedding from the high-dimensional sample data and fix the distortions at each point with estimated Riemannian geodesic distance and volume density function. This estimator could be further applied for unsupervised tasks such as anomaly detection where the outliers are the lowest density points.

For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][]{Heidenreich2013-bl}[ for details]{Scott2015-vl}.
For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][]{Jones1992-ef,Cao1994-st,Jones1996-cb}[ for reviews]{Wand1994-xu}. The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}. In this paper, we focus on the multivariate kernel density estimation.

Note that a fixed bandwidth matrix \(\pmb{H}\) is a global smoothing parameter for all data points. However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
The bandwidth is varied depending on either the location of the sample points {[}sample smoothing estimator; \textcite{Terrell1992-ut}{]} or that of the estimated points {[}balloon estimator; \textcite{Terrell1992-ut}{]}. In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the \emph{variable/adaptive kernel density estimation} {[}VKDE; Section 6.6 of \textcite{Scott2015-vl}{]} unless otherwise stated.
However, these kernel density estimators are based on random samples in the Euclidean space.

@@

\hypertarget{s1simulation}{%
\section{\texorpdfstring{Appendix: Simulation of \(S^1\) for manifold density estimation}{Appendix: Simulation of S\^{}1 for manifold density estimation}}\label{s1simulation}}

To start with, we demonstrate the use of the proposed DC-KDE on a \(1\)-dimensional example. For a \(S^1\) curve centered at \(O\) with radius \(1\), the coordinated of points on the curve could be expressed in polar coordinates \((1, \alpha)\) or Cartesian coordinates \((x, \sqrt{1-x^2})\).

The first simulation starts with a \(1\)-dimensional example.

Manifold: \(S^1\) with center 0 and radius \(r=1\)

Polar coordinates: \((r, \theta)\)

Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with \(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).

Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and \(y = \sin(\theta) \in [0, 1]\)

Density of \(\theta\): \(f(\theta) = 2 / \pi\)

Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)

The CDF of X is given by
\[
P(X \leq x) = P(\cos(\theta) \leq x) \\
= P(\theta \geq \arccos(x)) \\
= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
= 1 - \frac{2}{\pi} \arccos(x).
\]
Then the density of X is the derivative of the CDF w.r.t. \(x\) \[
f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
= \frac{2}{\pi \sqrt{1 - x^2}}.
\]

Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could
estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\)
and compare it with its true density \(f(x)\).
We use the density estimator in \textcite{Pelletier2005-vu}. For point \(p \in S^1\), the density at \(p\) is estimated by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
\]
where \(d=1\), \(r\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).

Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\) with \(r=1\). Therefore, the geodesic distance is
\[
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
\]
As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).

{[}When using normal coordinates(orthogonal basis) for tangent vector in \(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}

The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
\[
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
\]
is smooth in a local neighborhood \(U\) of \(M\).

Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
\[
\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
\]
Therefore, the estimator for \(p \in S^1\) is given by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
\]
where the bandwidth \(r\) is selected with the plug-in bandwidth selector implemented by the R package \emph{ks}.

\hypertarget{how-to-derive-volume-density-function-theta_x_ip}{%
\subsubsection{\texorpdfstring{How to derive volume density function \(\theta_{X_i}(p)\)}{How to derive volume density function \textbackslash theta\_\{X\_i\}(p)}}\label{how-to-derive-volume-density-function-theta_x_ip}}

\hypertarget{example-of-s2-in-henry2009-om.}{%
\paragraph{\texorpdfstring{Example of \(S^2\) in \textcite{Henry2009-om}.}{Example of S\^{}2 in @Henry2009-om.}}\label{example-of-s2-in-henry2009-om.}}

Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\). Let \(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\) is the orthonormal basis of \(\mathbb{R}^3\). Consider the exponential chart \((U, \Psi)\) induced by the parametrisation \(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by
\[
\Psi^{-1}(s,t) = \cos(r)p + \frac{\sin(r)}{r}(sv+tw),
\]
if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where \(r=\sqrt{s^2 + t^2}\).
Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
Then we take the partial derivatives of \(\Psi^{-1}(s,t)\) with respect to the basis of the tangent space, \(s\) and \(t\), and have that
\[
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
\]
\[
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
\]
The coefficients for \(v\) and \(w\) gives the local coordinates of two tangent vectors.
Further, since the riemannian metric is defined as the inner product between two tangent vectors, each component of the riemannian metric could be calculated as \(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\).
Therefore, we obtain that the volume density on the sphere is
\[
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
\]
for \(q \neq p, -p\), and \(\theta_p(p) = 1\).

\hypertarget{s1}{%
\paragraph{\texorpdfstring{\(S^1\)}{S\^{}1}}\label{s1}}

Similarly in \(S^1\), we could let \(\{v, p\}\) be the orthonormal basis of \(\mathbb{R}^2\). Then the exponential chart is induced by the parametrization
\[
\Psi^{-1}(s) = \cos(r)p + \frac{\sin(r)}{r}sv,
\]
and \(\Psi^{-1}(0) = p\) where \(r = s\). Again \(r = d_g(p,q)\) if \(q = \exp_p(sv)\). Then we have
\[
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = - \sin(r)p + \cos(r)v.
\]
So the riemannian metric is \(g_q = \cos(r)\) and further,
\[
\theta_p(q) = \sqrt{|\cos(r)|} = \sqrt{|\cos(d_g(p,q))|},
\]
where the geodesic distance is \(d_g(p,q) = \arccos(p) - \arccos(q)\), and \(\theta_p(p) = 1\).

>>>>

\author{Fan~Cheng, Anastasios~Panagiotelis, Rob J~Hyndman}
\addresses{\textbf{Fan Cheng}\newline
Monash University
\newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
\textbf{Anastasios Panagiotelis}\newline
University of Sydney
\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
\textbf{Rob J Hyndman}\newline
Monash University
\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
}

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of \(\mathbb{R}^d\). Applications of kernel density estimation \autocites[KDE;][]{Parzen1962-gt}{Chen2017-dw} include finding hot spots of traffic network in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac}, prime prediction via Twitter messages \autocite{Gerber2014-tq}, and so on. However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than \(\mathbb{R}^d\), kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted. \textcite{Pelletier2005-vu} propose a kernel density estimator based on the Riemannian geodesic distance of the manifold but it is only applicable when the underlying manifold is known. Manifold learning algorithms could be applied to reduce the dimension and get an approximation of the manifold, but different manifold learning algorithms could induce different distortions of the same manifold. Therefore, we propose a distortion-corrected kernel density estimator for Riemannian manifolds embedded in more than \(\mathbb{R}^d\). Our estimator could be applied to any reasonable manifold learning embedding from the high-dimensional sample data and fix the distortions at each point with estimated Riemannian geodesic distance and volume density function. This estimator could be further applied for unsupervised tasks such as anomaly detection where the outliers are the lowest density points.

For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE \autocite{Jones1990-oe,Terrell1992-ut}.
Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation \autocite{Jones1992-ta,Sain1994-gr} and plug-in methods \autocites[See][]{Heidenreich2013-bl}[ for details]{Scott2015-vl}.
For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated \autocites[See][]{Jones1992-ef}{Cao1994-st}{Jones1996-cb}[ for reviews]{Wand1994-xu}. The generalization to multivariate case could mostly be found in \textcite{Duong2003-sp}, \textcite{Duong2004-rh}, and \textcite{Chacon2010-wm}. In this paper, we focus on the multivariate kernel density estimation.

Note that a fixed bandwidth matrix \(\pmb{H}\) is a global smoothing parameter for all data points. However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed.
The bandwidth is varied depending on either the location of the sample points \autocite[sample smoothing estimator;][]{Terrell1992-ut} or that of the estimated points \autocite[balloon estimator;][]{Terrell1992-ut}. In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the \emph{variable/adaptive kernel density estimation} \autocite[VKDE; Section 6.6 of][]{Scott2015-vl} unless otherwise stated.
However, these kernel density estimators are based on random samples in the Euclidean space.

@@
